---
title: "Modeling Assignment: Swire Capstone"
author: "Ethan Aslami, Riley Walburger, Tara Connin"
date: 10/26/2025
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    number-depth: 2
    toc-location: left
    toc-title: "Contents"
    code-fold: false
    code-summary: "🖱️ Click to Show Code"
    embed-resources: true
execute:
  include: true
  eval: true    
  warning: false
  message: false---
---

# Introduction 
The goal of this modeling notebook is to develop a few predictive models that estimate the likelihood of cart abandonment within the MyCoke360 platform. The model will use features created from customer behavior, session activity, timing relative to cutoff windows, and operational scheduling to identify factors associated with incomplete purchases. By predicting the probability that a customer will abandon their cart, the analysis supports Swire Coca-Cola’s objective of improving online order completion and improving digital sales performance.

This notebook focuses on implementing a regression-based modeling approach to establish a strong baseline for prediction and interpretability. The process includes conducting feature engineering, defining a baseline model, selecting appropriate models, and evaluating models using cross-validation. Model performance will be assessed using the area under the ROC curve. The best-performing model will be validated for reliability and used to identify key factors associated with cart abandonment.

# Setup + Data Prep
```{r}
library(data.table)
library(caret)
library(tidyverse)
library(janitor)
library(pROC)
library(randomForest)
library(xgboost)
library(ranger)
library(smotefamily)


# data with abandonment 
order_window_final <- fread("order_window_with_abandonment.csv") 

material <- fread("data/material.csv") |> 
  clean_names()


orders <- fread("data/orders.csv") |> 
  clean_names()

sales <- fread("data/sales.csv") |> 
  clean_names()

customer <- fread("data/customer.csv") |> 
  clean_names() |> 
  rename(customer_id = customer_number, # match google analytics
         plant_id = sales_office)

google_analytics <- fread("data/google_analytics.csv") |> clean_names()



glimpse(order_window_final)
glimpse(material)
glimpse(orders)
glimpse(sales)
glimpse(customer)
```


### Basic data understanding and target inspection
```{r}
# Check target distribution
target_summary <- order_window_final[, .(
  total_obs = .N,
  abandoned = sum(cart_abandonment_flag, na.rm = TRUE),
  pct_abandoned = mean(cart_abandonment_flag, na.rm = TRUE) * 100
)]

# Check unique customers and order windows
counts_summary <- order_window_final[, .(
  unique_customers = uniqueN(customer_id),
  unique_order_windows = uniqueN(order_window_id)
)]

# Report: target_summary, na_summary (top 10 rows), counts_summary
target_summary
counts_summary

# Plot abandoned vs non-abandoned counts
ggplot(order_window_final, aes(x = factor(cart_abandonment_flag))) +
  geom_bar(fill = "darkgreen") +
  scale_x_discrete(labels = c("Not Abandoned", "Abandoned")) +
  labs(
    title = "Cart Abandonment Distribution",
    x = "Cart Abandonment",
    y = "Count"
  ) +
  theme_minimal()

```

- This dataset has a very high target imbalance with a very small percentage of the data containing abandoned carts 



## Feature Engineering 
- Here we add features from the customer data, sales data, orders data, and google analytics data aggregated at the customer level
```{r, results='hide'}

# --- CUSTOMER INFO MERGE ---
features_v1 <- merge(
  order_window_final,
  customer[, .(customer_id,
               sales_office_description,
               distribution_mode_description,
               shipping_conditions_description,
               cold_drink_channel_description,
               customer_sub_trade_channel_description)],
  by = "customer_id",
  all.x = TRUE
)

# --- RECENT ORDER ACTIVITY ---
# Compute order frequency and recency at customer level
orders_summary <- orders[, .(
  total_orders = .N,
  avg_order_qty = mean(order_quantity, na.rm = TRUE),
  days_since_last_order = as.numeric(Sys.Date() - max(created_date_est, na.rm = TRUE))
), by = customer_id]

# --- RECENT SALES ACTIVITY ---
sales_summary <- sales[, .(
  total_sales_tx = .N,
  total_volume = sum(physical_volume, na.rm = TRUE),
  avg_profit = mean(gross_profit_dead_net, na.rm = TRUE)
), by = customer_id]

# --- MERGE EVERYTHING ---
features_v1 <- merge(features_v1, orders_summary, by = "customer_id", all.x = TRUE)
features_v1 <- merge(features_v1, sales_summary, by = "customer_id", all.x = TRUE)

# --- HANDLE NAs ---
features_v1[is.na(total_orders), total_orders := 0]
features_v1[is.na(total_sales_tx), total_sales_tx := 0]

# Report: head(features_v1), summary of numeric columns (e.g., summary(select_if(is.numeric)))
head(features_v1)
summary(features_v1[, .(total_orders, avg_order_qty, days_since_last_order, total_sales_tx, total_volume, avg_profit)])
```


```{r, results='hide'}
# Step 3 - Feature Engineering

features_v2 <- copy(features_v1)

# --- TIME FEATURES ---
features_v2[, order_month := month(order_window_start)]
features_v2[, order_weekday := weekdays(order_window_start)]
features_v2[, order_year := year(order_window_start)]

# --- RECENCY FLAG ---
features_v2[, recent_order_flag := fifelse(days_since_last_order <= 30, 1, 0)]

# --- BEHAVIORAL RATIOS ---
features_v2[, avg_profit_per_tx := avg_profit / total_sales_tx]
features_v2[, volume_per_order := total_volume / total_orders]
features_v2[, order_freq_ratio := total_orders / total_sales_tx]

# --- HANDLE INF / NA ---
num_cols <- names(features_v2)[sapply(features_v2, is.numeric)]
for (col in num_cols) {
  features_v2[[col]][is.infinite(features_v2[[col]])] <- NA
  features_v2[[col]][is.na(features_v2[[col]])] <- median(features_v2[[col]], na.rm = TRUE)
}

# --- ENCODE CATEGORICALS ---
cat_cols <- c(
  "sales_office_description",
  "distribution_mode_description",
  "shipping_conditions_description",
  "cold_drink_channel_description",
  "customer_sub_trade_channel_description"
)

# Keep only columns that exist in features_v2
cat_cols <- cat_cols[cat_cols %in% names(features_v2)]

# Only convert if there are columns left
if (length(cat_cols) > 0) {
  features_v2[, (cat_cols) := lapply(.SD, as.factor), .SDcols = cat_cols]
}


# --- AGGREGATED CUSTOMER FEATURES FROM GOOGLE ANALYTICS ---
customer_features <- google_analytics[, .(
  total_events = .N,                                         # Total number of events for this customer
  total_purchases = sum(event_name == "purchase", na.rm = TRUE),  # Count of purchase events
  total_views = sum(grepl("view", event_name), na.rm = TRUE),     # Count of all view-type events (page views, item views)
  active_days = uniqueN(event_date),                         # Number of unique days customer was active
  mobile_events = sum(device_category == "mobile", na.rm = TRUE), # Number of events from mobile devices
  desktop_events = sum(device_category == "desktop", na.rm = TRUE), # Number of events from desktop devices
  top_os = names(sort(table(device_operating_system), decreasing = TRUE))[1], # Most frequently used OS
  distinct_pages = uniqueN(event_page_name[event_page_name != "null"]), # Count of unique pages visited
  distinct_items = uniqueN(unlist(str_split(items[items != "[]"], ","))) # Count of unique items interacted with
), by = customer_id]

# Merge GA features into features_v2
features_v2 <- merge(features_v2, customer_features, by = "customer_id", all.x = TRUE)

# Handle NAs from merge
num_cols_ga <- setdiff(names(customer_features), "customer_id")
for (col in num_cols_ga) {
  if (is.numeric(features_v2[[col]])) {
    features_v2[[col]][is.na(features_v2[[col]])] <- 0
  } else {
    features_v2[[col]][is.na(features_v2[[col]])] <- "Unknown"
  }
}

# Convert top_os to factor
features_v2[, top_os := as.factor(top_os)]

# --- RATIOS FROM GA ---
features_v2[, purchase_rate := fifelse(total_views > 0, total_purchases / total_views, 0)]
features_v2[, mobile_ratio := fifelse(total_events > 0, mobile_events / total_events, 0)]

# --- FINAL MODEL DATASET ---
model_data <- features_v2[, .(
  customer_id, order_window_id, cart_abandonment_flag,
  
  # Order / sales features
  total_orders, total_sales_tx, avg_profit, total_volume, avg_order_qty,
  days_since_last_order, recent_order_flag,
  avg_profit_per_tx, volume_per_order, order_freq_ratio,
  order_month, order_weekday,
  sales_office_description,
  shipping_conditions_description, cold_drink_channel_description,
  customer_sub_trade_channel_description,
  
  # GA aggregated features
  total_events, total_purchases, total_views, active_days,
  mobile_events, desktop_events,
  top_os, distinct_pages, distinct_items,
  
  # GA derived ratios
  purchase_rate, mobile_ratio
)]

# --- CHECK RESULTS ---
str(model_data)
summary(model_data$cart_abandonment_flag)
head(model_data)

```

# Modeling 
- We will use 3 different models for this section 
    - Logistic Regression 
    - Random Forest Model 
    - XGBoost Model
- For each model we will go through the process of setting it up then we will show the model results


## Set up data partition 
- Here we create an 80/20 Train/Test split and we downsample the data so that we can run the models more efficiently for this assignment. 
```{r}
set.seed(123)
# Setup a 80-20 split for train/test split 
# Train/test split
train_idx <- createDataPartition(model_data$cart_abandonment_flag, p = 0.8, list = FALSE)
train_full <- model_data[train_idx]
test_data  <- model_data[-train_idx]

# Drop IDs / constants first (before SMOTE)
cols_to_drop <- c("customer_id", "order_window_id")
train_full <- train_full[, !cols_to_drop, with = FALSE]
test_data  <- test_data[, !cols_to_drop, with = FALSE]

# Ensure factor target
train_full[, cart_abandonment_flag := factor(cart_abandonment_flag, levels = c(0, 1))]
test_data[, cart_abandonment_flag  := factor(cart_abandonment_flag,  levels = c(0, 1))]

# THis is for SMOTE but the models run too slowly with this so we couldn't run it 

# # Separate predictors and target
# x_train <- train_full[, !"cart_abandonment_flag", with = FALSE]
# y_train <- as.numeric(as.character(train_full$cart_abandonment_flag))

# # Identify categorical columns
# cat_cols <- names(x_train)[sapply(x_train, is.factor) | sapply(x_train, is.character)]

# # Convert to dummies using caret
# if(length(cat_cols) > 0){
#   dummies <- dummyVars(~ ., data = x_train, fullRank = TRUE)
#   x_train <- as.data.table(predict(dummies, newdata = x_train))
# }

# # Now all columns are numeric, ready for SMOTE
# set.seed(123)
# smote_out <- SMOTE(X = x_train, target = y_train, K = 5, dup_size = 10)

# # Combine back
# train_data <- as.data.table(cbind(smote_out$data, cart_abandonment_flag = smote_out$data$class))
# train_data[, class := NULL]  # remove duplicate
# train_data[, cart_abandonment_flag := factor(cart_abandonment_flag, levels = c(0, 1))]


# caret's downSample() expects predictors and a factor outcome
# We use this downsample function and lose a lot of data because of the target imbalance but 
# it is the only thing we can do right now to run the model efficiently 
train_data <- downSample(
  x = subset(train_full, select = -cart_abandonment_flag),
  y = as.factor(train_full$cart_abandonment_flag),
  yname = "cart_abandonment_flag"
)

train_data <- as.data.table(train_data)


# Check new balance
table(train_data$cart_abandonment_flag)
prop.table(table(train_data$cart_abandonment_flag))
```


## Baseline Logistic Regression 
- We will use a baseline model using only the numeric predictors for simplicity. 

## Modeling Process

```{r}
# Keep only numeric predictors
numeric_cols <- names(train_data)[sapply(train_data, is.numeric)]
# Include the target column
numeric_cols <- c("cart_abandonment_flag", numeric_cols[numeric_cols != "cart_abandonment_flag"])
train_data <- train_data[, ..numeric_cols]

# Ensure target is numeric 0/1
train_data[, cart_abandonment_flag := as.numeric(as.character(cart_abandonment_flag))]

# Fit baseline logistic regression
logit_model <- glm(cart_abandonment_flag ~ ., 
                   data = train_data, 
                   family = binomial)

# Summary for interpretability
summary(logit_model)

# Predicted probabilities
preds_logit <- predict(logit_model, type = "response")

```

### Modeling Results 

```{r}
# See results 
# coefficients
coef_df <- data.frame(
  Feature = names(coef(logit_model))[-1],
  Coefficient = coef(logit_model)[-1]
)
coef_df[order(abs(coef_df$Coefficient), decreasing = TRUE), ]


# Predict on test data
preds_test <- predict(logit_model, newdata = test_data, type = "response")

# Evaluate AUC on test data
auc_test <- auc(test_data$cart_abandonment_flag, preds_test)
cat("Baseline Logistic Regression Test AUC:", round(auc_test, 4), "\n")


```

## Random Forest 
- These next 2 models will use all the variables categorical and numeric for prediction

### Modeling Process
```{r}

# Run random forest
rf_model <- ranger(
  cart_abandonment_flag ~ .,
  data = train_data,
  num.trees = 200,
  probability = TRUE,
  importance = "impurity_corrected",
  sample.fraction = 0.5
)

# Predict on test set
rf_preds <- predict(rf_model, data = test_data)$predictions[,2]


```

### Modeling Performance 
```{r}
# See feature importance 
imp <- sort(rf_model$variable.importance, decreasing = TRUE)

# Horizontal barplot of feature importance
par(mar = c(5, 12, 4, 2))  # increase left margin
barplot(
  rev(head(imp, 15)),
  names.arg = rev(names(head(imp, 15))),
  horiz = TRUE,
  las = 1,             # horizontal text
  col = "steelblue",
  main = "Top 15 Feature Importances",
  cex.names = 0.8      # shrink label size if needed
)

# See results
auc(test_data$cart_abandonment_flag, rf_preds)

```


## XGBoost 

### Modeling Process

```{r, results='hide'}
# Prepare matrix
target_train <- as.numeric(as.character(train_data$cart_abandonment_flag))
features_train <- model.matrix(cart_abandonment_flag ~ . - 1, data = train_data)
dtrain <- xgb.DMatrix(data = features_train, label = target_train)

target_test <- as.numeric(as.character(test_data$cart_abandonment_flag))
features_test <- model.matrix(cart_abandonment_flag ~ . - 1, data = test_data)
dtest <- xgb.DMatrix(data = features_test, label = target_test)

# Cross-validation
params <- list(
  objective = "binary:logistic",
  eval_metric = "auc",
  eta = 0.1,
  max_depth = 6,
  subsample = 0.7,
  colsample_bytree = 0.7
)

set.seed(123)
cv_results <- xgb.cv(
  params = params,
  data = dtrain,
  nrounds = 100,
  nfold = 5,
  early_stopping_rounds = 5,
  maximize = TRUE,
  verbose = 1
)

```

```{r}
# Define the Best iteration
best_iter <- cv_results$best_iteration

# Train final model using the best iteration
xgb_model <- xgb.train(
  params = params,
  data = dtrain,
  nrounds = best_iter
)

```

### Modeling Performance 

```{r}
# Feature importance
importance <- xgb.importance(model = xgb_model)
xgb.plot.importance(importance_matrix = importance, top_n = 15)


# Columns used to train
train_cols <- colnames(dtrain)

# Reorder test columns to match training
dtest <- xgb.DMatrix(data = as.matrix(test_data[, ..train_cols]), 
                     label = test_data$cart_abandonment_flag)

# Predict on test
xgb_preds <- predict(xgb_model, dtest)
auc(target_test, xgb_preds)

```


# Results 

Using three different models, we’ve started to get a better sense of the factors that influence cart abandonment. While we’ll continue iterating and refining the models before the presentation, our initial results already show some interesting patterns. Total orders — the total number of orders a customer has made over their history — seems to affect abandonment. Total events also stands out, with customers who were very active on the platform. Days since last order is another important factor, as longer gaps between orders seem to contribute to abandonment. The number of distinct items a customer interacts with also looks relevant. We also think it would be interesting to dig further into mobile vs desktop events to see how device type impacts abandonment. Overall, this step gives us a solid starting point for modeling, and we’ll keep improving it as we prepare for the presentation. We still think that the main outcome of this part of the project shoule be to provide actionable insights for Swire and we will keep working on ways to define these reccomendations throughout the next couple weeks. 

